Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-12 paper-filmat.tex
More on models of set theory

\begin{ex}The \emph{standard topos}~$\Set$ contains all
sets as its objects and all maps between sets as morphisms.\end{ex}

@@

\subsection{Relation to models of set theory} In set theory, philosophy and
logic, models of set theories ... XXX

Toposes can be pictured as
``universes in which we can do mathematics'' in much the same way as models of
set theory. Just as using forcing and other techniques we can construct new
models of set theory from given ones, thereby exploring the set-theoretic
multiverse, we can construct new toposes from given toposes.

However, there are important differences between the notion of mathematical
universe as provided by toposes and as provided by models of set theory, both
regarding the subject matter and the reasons for why we are interested in them.

@@

In fact,
to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
is a topos~$\Set_M$ such that a statement holds in the internal language
of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the things
which~$M$ believes to be sets, and its morphisms are those things which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
equivalent, their associated toposes will not be equivalent.} This
includes~$V$, the trivial inner model consisting of all sets there are, in case
such a thing exists in one's chosen ontology; in this case the associated
topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
if and only if it's true in the ordinary mathematical sense.

@@

\subsection{Variants of the \effective topos} The \effective topos belongs to a
wider class of \emph{realizability toposes}. These can be obtained by repeating
the construction of the \effective topos with any other reasonable model of
computation in place of Turing machines. The resulting toposes will in general
not be equivalent and reflect higher-order properties of the employed models.
Two of these further toposes are of special philosphical interest.

\paragraph{Hypercomputation}
Firstly, in place of ordinary Turing machines, one can employ the
\emph{infinite-time Turing machines} pioneered by Hamkins and
Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
in that they can run for ``longer than infinity''; more precisely, the
computational steps are indexed by the ordinal numbers instead of the natural
numbers. For instance, an infinite-time Turing machine can trivially decide the
twin prime conjecture, by simply walking along the natural number line and
recording any twin primes it finds. Then, on day~$\omega$, it can observe
whether it has found finitely many twins or not.

@@

\paragraph{Machines of the real physical world} A second variant of the
\effective topos is obtained by using machines of the real physical world
instead of abstract Turing machines in its construction. In doing so, we of
course leave the realm of mathematics, as real-world machines are not objects
of mathematical studies, but still it is interesting to see which commitments
about the natural of the physical world imply which internal statements of the
resulting topos.

For instance, Andrej Bauer proved that inside this topos any function~$\RR \to
\RR$ is continuous if, in the physical world, only finitely many computational
steps can be carried out in finite time and if it is possible to form
tamper-free private communication channels~\cite{XXX}.  % XXX: look up precise formulation

@@

\subsection{Real numbers}
As detailled in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
actually a Turing machine computing arbitrarily-good consistent rational
approximations. A similarly drastic shift in meaning, though in an orthogonal
direction, occurs with~$\Sh(X)$. What~$\Sh(X)$ believes to be a (Dedekind) real
number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
function~$a : X \to \RR$.

>>>>

\begin{defn}The \emph{standard topos}~$\Set$ is the category which contains all
sets as its objects and all maps between sets as morphisms.\end{defn}

@@

\subsection{Relation to models of set theory} In set theory, philosophy and
logic, models of set theories are studied. These are structures~$(M,\in)$
validating the axioms of some set theory such as Zermelo--Fraenkel set theory
with choice~\textsc{zfc}, and they can be pictured as ``universes in which we
can do mathematics'' in much the same way as toposes.

In fact, to any model~$(M,\in)$ of a set theory such as~\textsc{zf}
or~\textsc{zfc}, there is a topos~$\Set_M$ such that a statement holds
in~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the entities
which~$M$ believes to be sets, and its morphisms are those entities which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
\textsc{etcs}~\cite{XXX}, and models are elementarily equivalent if their
associated toposes are equivalent as categories.}

\begin{ex}The topos~$\Set_V$ associated to the universe~$V$ of all sets (if
this structure is available in one's chosen ontology) coincides with the
standard topos~$\Set$.\end{ex}

In set theory, we use forcing and other techniques to construct new
models of set theory from given ones, thereby exploring the set-theoretic
multiverse. There are similar techniques available for constructing new toposes
from given ones, and some of these correspond to the techniques from set
theory.

However, there are also important differences between the notion of mathematical
universes as provided by toposes and as provided by models of set theory, both
regarding the subject matter and the reasons for why we are interested in them.

@@

@@

\subsection{Variants of the \effective topos} The \effective topos belongs to a
wider class of \emph{realizability toposes}. These can be obtained by repeating
the construction of the \effective topos with any other reasonable model of
computation in place of Turing machines. The resulting toposes will in general
not be equivalent and reflect higher-order properties of the employed models.
Two of these further toposes are of special philosophical interest.

\bigskip
\paragraph{Hypercomputation}
Firstly, in place of ordinary Turing machines, one can employ the
\emph{infinite-time Turing machines} pioneered by Hamkins and
Lewis~\cite{hamkins-lewis:ittm}. These machines model \emph{hypercomputation}
in that they can run for ``longer than infinity''; more precisely, the
computational steps are indexed by the ordinal numbers instead of the natural
numbers. For instance, an infinite-time Turing machine can trivially decide the
twin prime conjecture, by simply walking along the natural number line and
recording any twin primes it finds. Then, on day~$\omega$, it can observe
whether it has found finitely many twins or not.

@@

\bigskip
\paragraph{Machines of the real physical world} A second variant of the
\effective topos is obtained by using machines of the real physical world
instead of abstract Turing machines in its construction. In doing so, we of
course leave the realm of mathematics, as real-world machines are not objects
of mathematical studies, but still it is interesting to see which commitments
about the natural of the physical world imply which internal statements of the
resulting topos.

For instance, Andrej Bauer proved that inside this topos any function~$\RR \to
\RR$ is continuous if, in the physical world, only finitely many computational
steps can be carried out in finite time and if it is possible to form
tamper-free private communication channels~\cite{bauer:int-mathematics}.

@@

\subsection{Real numbers}
As detailled in Section~\ref{sect:eff-continuous}, what the \effective topos believes to be a real number is
actually a Turing machine computing arbitrarily-good consistent rational
approximations. A similarly drastic shift in meaning, though in an orthogonal
direction, occurs with~$\Sh(X)$. What~$\Sh(X)$ believes to be a (Dedekind) real
number~$a$ is actually a continuous family of real numbers on~$X$, that is, a
continuous function~$a : X \to \RR$.
